Type Ia Supernova Model

### 2015-06-26 doc/model.tex
presented to DES

\documentclass[preprint]{aastex}
\usepackage{amsmath,amssymb}
\usepackage{mathrsfs}
\usepackage{graphicx}
\newcommand{\mli}[1]{\mathit{#1}}
%\usepackage{epstopdf}

\bibliographystyle{plain}

@@

\section{Model}
Supernova cosmology analysis is based on the Hubble diagram: measurements of
redshift $z$ and distance modulus $\mu$.  The inferred Hubble diagram
can be expressed as the pdf of the true underlying
quantities given these measurements 
\begin{equation}
P(\vec{\mu},\vec{z}| \vec{\hat{\mu}}, \vec{\hat{z}}, \vec{\hat{D}}=1, \vec{\hat{T}}=\text{SN~Ia}),
\end{equation}
where $\hat{D}=1$ and $\hat{T}=\text{SN~Ia}$ indicate that the objects are discovered and typed
as SNe~Ia.  
The caret labels measurements and quantities derived
from measurements.  This Hubble diagram can then be used to infer parameters
of cosmological and dark energy models.

A model of the experiment is necessary to interpret the data, as in the
calculation of the likelihood function $P(\vec{\hat{\mu}}, \vec{\hat{z}}, \vec{\hat{D}}=1, \vec{\hat{T}}=\text{SN~Ia}|\vec{\mu},\vec{z})$.
A sketch of a such a model is shown in the Probabilistic Graphical Model
shown in Figure~\ref{pgm:fig}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.
   The bottom right corner containing the connection between observed light curves
   $\hat{f}$, transient spectral information $\hat{z}_S$, $\hat{\theta}_S$, and
   host information $\hat{z}_H$, $\hat{\theta}_H$ with the determined type, redshift, and distance
   modulus $\hat{\mathit{Type}}$, $\hat{z}$, and $\hat{\mu}$ is the purview of
   the traditional light-curve and Hubble diagram fitters.
   \label{pgm:fig}}
\end{figure}

@@

There is a set of redshifts
at which a transients explode $\vec{z}$ and for each (cosmological) redshift there is a corresponding distance
modulus $\vec{\mu}$; these $z$-$\mu$ pairs are the fundamental parameters that
specify the Hubble diagram.

Given that there is a transient at redshift $z$, it has a probability of being a particular
Type associated with a Host. 
The Host is anything that projects onto the line of sight of the transient and so contributes
to measurement noise.   For our purposes, the Type and Host include information on the
sources and line-of-sight effects that can be encapsulated as an effective SED. 
This SED and distance modulus determine the photon flux incident above the Earth's
atmosphere, $n$ and $n_g$.

The experiment has an optical chain summarized by the transmission functions $T(\lambda)$, which determine the expected counts of the transient and galaxy, $\mathit{ADU}$ and
$\mathit{ADU}_g$.  
The realized observed counts determine whether the transient
is discovered $\hat{D}$, and the 
calibration $\hat{Z}$ of the transmission functions set the
measured flux-calibrated light curves $\hat{f}$.

Spectra $\hat{\mathit{Spec.}}$ may be obtained for active candidates, from which typing, redshift and spectral parameters $\hat{z}_S$, $\hat{\theta}_S$ are derived.
From the detected galaxies Gals.\ a host is associate with the transient, the measured
host redshift and parameters are $\hat{z}_{H}$, $\hat{\theta}_H$.

Given the transient light curves, spectra, and host-galaxy measurements, the transient
type $\hat{\mathit{Type}}$, redshift $\hat{z}$, and distance modulus $\hat{\mu}$ are
determined.  Depending on what information is available, this is done by some combination
of a light-curve fitter, classifier, and empirical SN~Ia modeler; in this note this black box
is referred to as a Fitter.

@@

\begin{description}
\item[$\hat{D}$] Transient detection.  It depends on
the observed signal $\hat{f}$ and its noise, mostly through the signal-to-noise
$\mathit{SNR}$.
\item[$\hat{z}$] Cosmological redshift.  There is an ordering of how the redshift
is determined. If there is an unambiguous host its spectroscopic redshift $\hat{z}_H$ is
selected.  Next, the transient's spectroscopic redshift $\hat{z}_S$ is preferred.
Generally these spectroscopic redshifts have negligible uncertainties for our purposes.
When the host is ambiguous, has only a photometric redshift, and/or there is
no transient redshift,
the calibrated supernova light curve $\hat{f}$ and all other information is combined to
infer redshift.
\item[$\hat{\mathit{Type}}$] Transient classification.  Traditionally only objects
typed as SN~Ia are used in the analysis: alternatively a probability of being
SN~Ia could be used.  The soundest classification comes from a transient
spectrum $\hat{\mathit{Spec}}$.  Otherwise typing is informed by host information
$\hat{z}_H$ and $\hat{\theta}_H$, and
calibrated light curves $\hat{f}$.
\item[ $\hat{\mu}$]  Distance modulus.  This is typically done with a light-curve
analysis using $\hat{f}$, $\hat{z}_S$ or
$\hat{z}_H$, and transient spectral features $\hat{\theta}_S$ and host
properties $\hat{\theta}_H$.
\end{description}

Generically the typing, redshift, and distance modulus are obtained
using a Fitter, with the details depending on the available information.
A calibration of the empirical SN~Ia absolute magnitude model is fit in this step.
Therefore $\hat{z}$, $\hat{\mathit{Type}}$
and $\hat{\mu}$ may be correlated, and correlations between transients are expected.

\begin{description}
\item[$\hat{z}_H$, $\hat{\theta}_H$] Host galaxy redshift and properties from spectroscopic
and photometric data, based on association
of the host from the pool of detected galaxies Gals.
\item[Gals] Detected galaxies from which a host is associated with a transient.
\item[$\hat{z}_S$, $\hat{\theta}_S$] Transient redshift and properties from
spectroscopic data $\hat{\mathit{Spec.}}$
\item[ $\hat{\mathit{Spec.}}$] Spectrum of the transient. I skip the details of the actual measurement  make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
\item[$\hat{f}$] Measured light curves.  Depends on the expected counts of
the transient and background $\mathit{ADU}$, $\mathit{ADU}_g$
(as a contributor to noise), and the photometric zeropoints $\hat{Z}$.
\item[$\hat{\mathit{ADU}}$, $\hat{\mathit{ADU}_g}$]  Expected counts
of the transient and background galaxy respectively.  Depends
on the incident fluxes $n$, $n_g$ and the instrumental
transmission $T(\lambda)$.
\item[$\hat{Z}$] Calibration zeropoints.  Depends on the calibration process
and the underlying transmission functions $T(\lambda)$.  This may be a zeropoint
but most probably will be calibratied transmission functions.
\item[$T(\lambda)$] DES transmission functions.
\item[$n$, $n_g$] The incident fluxes at Earth of the transient and background
galaxy.  Depends on  $z$ and $\mu$, and the SEDs of the Type and Hiost.
\item[Type/Host]  The type of transient and projected galaxies (referred to as host),
used to the specify the
underlying SED.  The realization depends on the redshift $z$.
\item[$\vec{\mu}$, $\vec{z}$]  The true distance moduli and cosmological redshifts under
consideration.  Each redshift has a unique distance modulus.
\end{description}

\section{Comments on Parts of the Model}

\subsection{Type}
The model allows for transients of all types to be detected
and potentially classified as SN~Ia. 
The full population of potential interlopers and their rates, $P(\text{Type} | z)$,
is poorly constrained.
I therefore propose that the model be constrained to SNe~Ia only,
and have the experiment measure
$P(non-Ia |  \hat{\mathit{Type}}=\text{SN~Ia})$,
$\left. \langle \hat{\mu}-\mu(z)\rangle \right|_{non-Ia}$ and its uncertainty.
This would be added as a correction to the Hubble diagram under the assumption
that all objects are SNe~Ia.

\subsection{The Fitter}
The model contains the term
\begin{align}
\begin{split}
P(\hat{\mu},\hat{z}, \hat{D}=1, \hat{\mathit{Type}}=\text{SN~Ia}|X) & =  P(\hat{D=1}|X) P(\hat{\mu},\hat{z},\hat{\mathit{Type}}=\text{SN~Ia}|X)\\
&=  P(\hat{D=1}|\hat{f}) P(\hat{\mu},\hat{z}, \hat{\mathit{Type}}=\text{SN~Ia}|X)\\
\end{split}
\label{fitter:eqn}
\end{align}
where $X=\{\hat{f}, \hat{\mathit{Spec.}}, \hat{z}_S, \hat{\theta}_S, \hat{z}_H, \hat{\theta}_H\}$
represents relevant measurements.  The first term on the right side of Eqn.~\ref{fitter:eqn}
is the detection efficiency.  The second term is what I have generically called the
Fitter.

Traditional supernova cosmology analysis is covered by the arrows that lead to
the measured type, redshift, and distance modulus.  Within this larger
context, probably the supernova absolute magnitudes should be calibrated relative
to a model that does not depend on the cosmological parameters.

Need to calibrate $P(x|y,\hat{T=\text{SN~Ia}})$.

>>>>

\documentclass[preprint]{aastex}
\usepackage{amsmath,amssymb}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{natbib}
\bibliographystyle{apj}

\newcommand{\mli}[1]{\mathit{#1}}
%\usepackage{epstopdf}

@@

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
classified as Type~Ia supernovae (SNe~Ia).  The inferred Hubble diagram can be
expressed as
\begin{equation}
P(\vec{\mu},\vec{z}| \vec{\hat{D}}=1, \vec{\hat{ADU}}, \vec{\hat{T}}_S=\text{SN~Ia},\vec{\hat{z}}_S,
\vec{\hat{\theta}}_S,\vec{\hat{z}}_H,\vec{\hat{\theta}}_H),
\label{hd:eqn}
\end{equation}
where the inferred quantities are the set of distance moduli ($\vec{\mu}$)
and redshifts ($\vec{z}$), and 
the measurements include: $\hat{D}$ the boolean that indicates that
the transient was discovered; $\hat{ADU}$ the transient photometry;
spectroscopically derived transient
typing, redshift and parameters $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$;
and the redshift and parameters of the associated host galaxy $\hat{z}_{H}$, $\hat{\theta}_H$.
The conditions $\vec{\hat{D}}=1$, $\vec{\hat{T}}_S,=\text{SN~Ia}$ hold as only transients
discovered and classified as SNe~Ia are modelled.  We use the convention of using carets $\hat{}$
to  label measurements.

A sketch of a model explaining the data is depicted through the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.
Although only transients that are detected and classified as
SN~Ia are modeled, the model accommodates the possibility that they could not have been found
or not classified as SN~Ia.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.
   The bottom right corner containing the connection between observed light curves
   $\hat{f}$, transient spectral information $\hat{z}_S$, $\hat{\theta}_S$, and
   host information $\hat{z}_H$, $\hat{\theta}_H$ with the determined type, redshift, and distance
   modulus $\hat{T}$, $\hat{z}$, and $\hat{\mu}$ is the purview of
   the traditional light-curve and Hubble diagram fitters.
   \label{pgm:fig}}
\end{figure}

One feature to note is that $\mu$ is a model parameter; most analyses instead
directly have cosmological parameters $\{\Omega_M, \Omega_\Lambda, w_0, w_a\}$.
A motivation is to provide results that can be applied to test arbitrary cosmological models:
indeed the model-dependence of the $\mu$'s reported for high-$z$ supernovae 
\citep[e.g.][]{2012ApJ...746...85S} is a source of confusion to the community.
An example of a non-parametric model for $\mu$ that is independent of cosmological
parameters is given in the Appendix of
\citet{2013ApJ...764..116W}.

@@

Discovered transients typed as SNe~Ia are modeled.  Each has
a cosmological redshift and corresponding distance modulus; these $z$-$\mu$ pairs are the  parameters that
constitute the Hubble diagram.  For the moment, the coordinates RA/Dec are considered
precisely determined from the images.

Each transient indexed by $i$ is of a certain type $T$ and comes from a  host galaxy of type $G$.  
(``Host'' includes the host as well as  all galaxies that project onto the coordinate of
the transient.)  The
transients (galaxies) are modeled and described by global parameters $\theta_T$ ($\theta_G$),
and  parameters specific to the transient $\theta_{Ti}$ ($\theta_{Gi}$).  For example, the SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would fill $\theta_{Ti}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would fill $\theta_T$.
The Type model includes  information on the
source and line-of-sight effects, which specify the effective SED. 
This SED, redshift, and  distance modulus determine the photon flux incident above the Earth's
atmosphere, $n$ and $n_g$.

The experiment has an optical chain summarized by the transmission functions $\phi(\lambda)$, which determine the expected counts of the transient and galaxy, $\mathit{ADU}$ and
$\mathit{ADU}_g$.   These expected counts in turn determine realized  transient counts 
$\mathit{ADU}_V$.
The signal to noise of the realized counts specify the probability of discovery
$\hat{D}$ .    The photometry of the subset of discovered transients
are denoted as $\hat{\mathit{ADU}}$.

For discovered transients, there may be spectra of the object itself
from which typing, redshift and spectral parameters $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$ are derived.
From amongst the detected galaxies $\hat{\mathit{Gals.}}$  a host is associated with the transient
based on its RA/Dec, its
host redshift (spectroscopic or photometric) and parameters are $\hat{z}_{H}$, $\hat{\theta}_H$.

The observatory transmission $\phi$ are calibrated $\hat{Z}$.

@@

The observables are:
\begin{description}
\item[$\hat{\mathit{Gals}}$] Detected galaxies from which a host is associated with a transient.
Depends on the model for galaxies $\theta_G$.
\item[$\hat{D}$] Transient detection.  It depends on
the realized signal $\hat{\mathit{ADU}_V}$ and its noise, mostly through the signal-to-noise
$\mathit{SNR}$.
\item[$\hat{\mathit{ADU}}$]  Photometry of discovered transients.  Exactly the value
of the  realized signal $\mathit{ADU}_V$ and discovery $\hat{D}$.
\item[$\hat{z}_H$, $\hat{\theta}_H$] Host galaxy redshift and properties from spectroscopic
and photometric data, based on association
of the host from the pool of detected galaxies $\hat{\mathit{Gals}}$ and the transient RA/Dec.
\item[$\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$] Transient redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
\item[$\hat{Z}$] Calibration zeropoints.  Depends on the calibration process
and the underlying transmission functions $\phi(\lambda)$.  This may be a zeropoint
but most probably will be calibrated transmission functions.
\end{description}

The following are elements of the model.
\begin{description}
\item[$\mathit{ADU}_V$]  Realized counts
of the transient.  Depends
on the expected counts of the transient and host $\mathit{ADU}$, $\mathit{ADU}_g$.
This is distinct from $\hat{\mathit{ADU}}$, which only has photometry of detected transients.
\item[$\mathit{ADU}$, $\mathit{ADU}_{g}$]  Expected counts
of the transient and background galaxy respectively.  Depends
on the incident fluxes $n$, $n_g$ and the instrumental
transmission $\phi(\lambda)$.
\item[$\phi(\lambda)$] DES transmission functions.
\item[$n$, $n_g$] The incident fluxes at Earth of the transient and background
galaxy.  Depends on  $z$ and $\mu$, and the SEDs of the Type and Host as
specified by $\theta_{Ti}$, $\theta_{Gi}$,$\theta_{T}$, $\theta_{G}$.
\item[$\theta_{Ti}$, $\theta_{Gi}$] Parameters that describe the properties of the individual transient
and host respectively.
\item[$\theta_{T}$, $\theta_{G}$] Parameters that describe the global properties of the transient
and host models.
\item[$T$,$G$]  The type of transient and projected galaxies (referred to as host), which
specifies which model to use. The realization depends on the redshift $z$.
\item[$\vec{\mu}$, $\vec{z}$]  The true distance moduli and cosmological redshifts under
consideration.  Each redshift has a unique distance modulus.
\item[$\mathbf{RA/Dec}$] The coordinate of the transient.  It will be the actual coordinate of the found discovered object.
\end{description}

\section{Subsections of the Model}

\subsection{Type}
\label{type:sec}
Given that our transient sample may not consist purely of SNe~Ia, the model
has a term of the form $P(T, G | z,\mu)$, which should account for all types of
objects that could potentially be mistaken for Type~Ia:
the rates and association with host galaxies are described for
the full population of potential interlopers.
%Note the condition $\hat{T}=\text{SN~Ia}$ means that non-Ia transients unlikely to
%be identified as SN~Ia can be described with poorer accuracy.
This term is not currently well known, particularly at the highest redshifts probed by DES.
It could be constructed through modeling, although then the uncertainties 
of that modeling must in turn be quantified.

I posit that a concise way of addressing this term is to only model $T=\text{SN~Ia}$,
and experimentally determine $P(\text{non-Ia})$, 
$\left. \langle \mu-\mu_{model} \rangle \right|_{non-Ia}$, and its uncertainty for the
sample of non-Ia's that are discovered and misclassified as SN~Ia.
This measurement is achieved through spectroscopic
typing of a subset drawn from the populations used in the Hubble diagram.
Other supernova surveys
with spectroscopic follow-up (e.g.\ SNLS) that replicate the DES population
can be used, for example on magnitude-limited samples with
complete  spectroscopic follow-up. 
The weighted bias would be added as a correction to the Hubble diagram under the assumption that all objects are SNe~Ia.

To minimize the contribution of non-Ia contamination we strive for
an excellent classifier so that $P(non-Ia) \rightarrow 0$.

\subsection{Host Matching}
\sloppy
The host-galaxy properties depends on the galaxy catalog and the transient
coordinates $P(\hat{z}_H, \hat{\theta}_H | \hat{\mathit{Gals.}}, \text{RA/Dec})$.  With a
spectrum of the transient and the underlying background, identification of the host galaxy
is usually straightforward.
Otherwise, projections or ambiguous cases can result in the misidentification of
the host.

A similar strategy can be taken as with non-Ia contamination in \S\ref{type:sec}:
analyze the supernova assuming a correct host match, and
experimentally determine $P(\text{mismatch} )$ and
$\left. \langle \hat{\mu}-\mu(z)\rangle \right|_{\text{mismatch}}$, $\left. \langle \hat{z}-z \rangle \right|_{\text{mismatch}}$ to correct bias in the Hubble diagram. 
We strive to have $P(\text{mismatch}) \rightarrow 0$.

\subsection{Type~Ia Supernova Model}
\label{snmodel:sec}
The model contains a piece $P(\hat{\mathit{ADU}}_{i} \hat{z}_{Si}| \mu_i,z_i,\theta_T, \theta_{Ti}, ,\phi(\lambda))$, which
is the piece that describes underlying SN~Ia properties and light-curve fits.

Need something more general?
$P(\hat{D}_i,\hat{\mathit{ADU}}_{i}, \hat{T}_{Si}, \hat{z}_{Si}, \hat{\theta}_{Si}, \hat{z}_{Hi}, \hat{\theta}_{Hi}
 | \mu_i,z_i,T_i,\theta_T,\theta_{Ti},\phi(\lambda))$
\subsection{Detection Efficiency}
There term $P(\hat{D}|\mathit{ADU}_V)$.  For all high-redshift supernova searches
I know of, this depends on the signal-to-noise of the data, and very weakly on the
structure of the underlying galaxy.  Our detection is now based on machine learning,
we should confirm that detection dependencies do not depend on our model parameters
(except for RA/Dec).

\section{Systematic Tests}
The Hubble diagrams of subsets of data quantify systematic uncertainty.  
Subsets that may be expected to show evidence for systematics include:
spectroscopically typed SNe~Ia; transients in deep (3) versus shallow (1/2) DES fields;
transients with NIR data; splits based on host properties.

A potentially more probative test is to compare the distance moduii inferred for the same
supernovae, but reducing the amount of data considered.  A subset of DES
transients have higher signal-to-noise, spectroscopic confirmation, and NIR coverage than
the average transient, we can take away that extra information in calculating distance modulus
to get $P(\mu | \mathit{data} - \mu | \mathit{data}^-, z | \mathit{data} - z | \mathit{data}^-)$.
It is possible that differences in $\mu$s and $z$s have smaller uncertainties that the
$\mu$ and $z$ alone due to common measurement uncertainties.

\bibliography{/Users/akim/Documents/alex}
